---
title: "Exercise 3-02"
author: "Riyanshi Bohra"
format: html
editor: visual
---

# Exercise 3: Classification: Alternative Techniques

# Setting Up

```{r}
# Installing necessary packages
if(!require(pacman))
  install.packages("pacman")

pacman::p_load(
  C50,                
  caret,              
  e1071,              
  keras,             
  kernlab,            
  lattice,            
  MASS,               
  mlbench,           
  nnet,               
  palmerpenguins,     
  party,              
  partykit,           
  randomForest,       
  rpart,              
  RWeka,              
  scales,            
  tidymodels,         
  tidyverse,        
  xgboost             
)
```

```{r}
# Show fewer digits
options(digits=3)
```

# Training and Test Data

```{r}
# Loading the Zoo dataset

data(Zoo, package="mlbench")
Zoo <- as.data.frame(Zoo)
Zoo |> glimpse()
```

```{r}
# Partitioning the data into 80% Train data and 20% test data

set.seed(123)  # for reproducibility
inTrain <- createDataPartition(y = Zoo$type, p = .8)[[1]]
Zoo_train <- dplyr::slice(Zoo, inTrain)
Zoo_test <- dplyr::slice(Zoo, -inTrain)
```

# Fitting Different Classification Models

```{r}
# Create a fixed sampling scheme (10-folds)
train_index <- createFolds(Zoo_train$type, k = 10)
```

## Conditional Inference Tree (Decision Tree)

```{r}
# Creating a ctree( Decision Tree)
ctreeFit <- Zoo_train |> train(type ~ .,
  method = "ctree",
  data = _,
    tuneLength = 5,
    trControl = trainControl(method = "cv", indexOut = train_index))
ctreeFit
```

```{r}
# Plotting the ctree
plot(ctreeFit$finalModel)
```

## C 4.5 Decision Tree

```{r}
C45Fit <- Zoo_train |> train(type ~ .,
  method = "J48",
  data = _,
    tuneLength = 5,
    trControl = trainControl(method = "cv", indexOut = train_index))
C45Fit
```

```{r}
C45Fit$finalModel
```

## K-Nearest Neighbors

```{r}
# Creatinga KNN Model
knnFit <- Zoo_train |> train(type ~ .,
  method = "knn",
  data = _,
  preProcess = "scale",
    tuneLength = 5,
  tuneGrid=data.frame(k = 1:10),
    trControl = trainControl(method = "cv", indexOut = train_index))
knnFit
```

```{r}
knnFit$finalModel
```

## PART (Rule-based classifier)

```{r}
# Creating a Rule-based classifier
rulesFit <- Zoo_train |> train(type ~ .,
  method = "PART",
  data = _,
  tuneLength = 5,
  trControl = trainControl(method = "cv", indexOut = train_index))
rulesFit
```

```{r}
rulesFit$finalModel
```

## Linear Support Vector Machines

```{r}
# Creating a Linear SVM model
svmFit <- Zoo_train |> train(type ~.,
  method = "svmLinear",
  data = _,
    tuneLength = 5,
    trControl = trainControl(method = "cv", indexOut = train_index))
svmFit
```

```{r}
svmFit$finalModel
```

## Random Forest

```{r}
# Create a random forest model
randomForestFit <- Zoo_train |> train(type ~ .,
  method = "rf",
  data = _,
    tuneLength = 5,
    trControl = trainControl(method = "cv", indexOut = train_index))
randomForestFit
```

```{r}
randomForestFit$finalModel
```

## Gradient Boosted Decision Trees (xgboost)

```{r}
# Creating a xgboost model 

xgboostFit <- Zoo_train |> train(type ~ .,
  method = "xgbTree",
  data = _,
  tuneLength = 5,
  trControl = trainControl(method = "cv", indexOut = train_index),
  tuneGrid = expand.grid(
    nrounds = 20,
    max_depth = 3,
    colsample_bytree = .6,
    eta = 0.1,
    gamma=0,
    min_child_weight = 1,
    subsample = .5
  ))
xgboostFit
```

```{r}
xgboostFit$finalModel
```

## Artificial Neural Network

```{r}
# Create an ANN Model
nnetFit <- Zoo_train |> train(type ~ .,
  method = "nnet",
  data = _,
    tuneLength = 5,
    trControl = trainControl(method = "cv", indexOut = train_index),
  trace = FALSE)
nnetFit
```

```{r}
nnetFit$finalModel
```

## Comparing Models

```{r}
# Collecting the performance metrics from the models trained on the same data

resamps <- resamples(list(
  ctree = ctreeFit,
  C45 = C45Fit,
  SVM = svmFit,
  KNN = knnFit,
  rules = rulesFit,
  randomForest = randomForestFit,
  xgboost = xgboostFit,
  NeuralNet = nnetFit
    ))
resamps
```

```{r}
# Calculate summary statistics

summary(resamps)
```

```{r}
# Plotting using lattice
library(lattice)
bwplot(resamps, layout = c(3, 1))
```

```{r}
# Perform inference about differences between models
difs <- diff(resamps)
difs
```

```{r}
# Finding summary statistics
summary(difs)
```

# Applying the Chosen Model to the Test Data

```{r}
# Choosing Random Forest Model for Test data
pr <- predict(randomForestFit, Zoo_test)
pr
```

```{r}
# Calculate the confusion matrix for the test data

confusionMatrix(pr, reference = Zoo_test$type)
```

# Comparing Decision Boundaries

```{r}
library(scales)
library(tidyverse)
library(ggplot2)
library(caret)

decisionplot <- function(model, data, class_var, 
  predict_type = c("class", "prob"), resolution = 3 * 72) {
  # resolution is set to 72 dpi if the image is rendered  3 inches wide. 
  
  y <- data |> pull(class_var)
  x <- data |> dplyr::select(-all_of(class_var))
  
  # resubstitution accuracy
  prediction <- predict(model, x, type = predict_type[1])
  # LDA returns a list
  if(is.list(prediction)) prediction <- prediction$class
  prediction <- factor(prediction, levels = levels(y))
  
  cm <- confusionMatrix(data = prediction, 
                        reference = y)
  acc <- cm$overall["Accuracy"]
  
  # evaluate model on a grid
  r <- sapply(x[, 1:2], range, na.rm = TRUE)
  xs <- seq(r[1,1], r[2,1], length.out = resolution)
  ys <- seq(r[1,2], r[2,2], length.out = resolution)
  g <- cbind(rep(xs, each = resolution), rep(ys, time = resolution))
  colnames(g) <- colnames(r)
  g <- as_tibble(g)
  
  ### guess how to get class labels from predict
  ### (unfortunately not very consistent between models)
  cl <- predict(model, g, type = predict_type[1])
  
  # LDA returns a list
  prob <- NULL
  if(is.list(cl)) { 
    prob <- cl$posterior
    cl <- cl$class
  } else
    if(!is.na(predict_type[2]))
      try(prob <- predict(model, g, type = predict_type[2]))
  
  # we visualize the difference in probability/score between the 
  # winning class and the second best class.
  # don't use probability if predict for the classifier does not support it.
  max_prob <- 1
  if(!is.null(prob))
    try({
      max_prob <- t(apply(prob, MARGIN = 1, sort, decreasing = TRUE))
      max_prob <- max_prob[,1] - max_prob[,2]
    }, silent = TRUE) 
  
  cl <- factor(cl, levels = levels(y))
  
  g <- g |> add_column(prediction = cl, probability = max_prob)
  
  ggplot(g, mapping = aes(
    x = .data[[colnames(g)[1]]], y = .data[[colnames(g)[2]]])) +
    geom_raster(mapping = aes(fill = prediction, alpha = probability)) +
    geom_contour(mapping = aes(z = as.numeric(prediction)), 
      bins = length(levels(cl)), linewidth = .5, color = "black") +
    geom_point(data = data, mapping =  aes(
      x = .data[[colnames(data)[1]]], 
      y = .data[[colnames(data)[2]]],
      shape = .data[[class_var]]), alpha = .7) + 
    scale_alpha_continuous(range = c(0,1), limits = c(0,1), guide = "none") +  
    labs(subtitle = paste("Training accuracy:", round(acc, 2))) +
     theme_minimal(base_size = 14)
}
```

# Penguins Dataset

```{r}
# We use two dimensions of the penguins dataset for visualization
set.seed(1000)
data("penguins")
penguins <- as_tibble(penguins) |>
  drop_na()

x <- penguins |> dplyr::select(bill_length_mm, bill_depth_mm, species)
x
```

```{r}
# Plot
ggplot(x, aes(x = bill_length_mm, y = bill_depth_mm, fill = species)) +  
  stat_density_2d(geom = "polygon", aes(alpha = after_stat(level))) +
  geom_point() +
  theme_minimal(base_size = 14) +
  labs(x = "Bill length (mm)",
       y = "Bill depth (mm)",
       fill = "Species",
       alpha = "Density")
```

## K-Nearest Neighbors Classifier

```{r}
# Visualize the KNN classifier with penguins dataset
# 1 neighbor
model <- x |> caret::knn3(species ~ ., data = _, k = 1)
decisionplot(model, x, class_var = "species") + 
  labs(title = "kNN (1 neighbor)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")
```

```{r}
# Visualize the KNN classifier with penguins dataset
# 3 neighbors
model <- x |> caret::knn3(species ~ ., data = _, k = 3)
decisionplot(model, x, class_var = "species") + 
  labs(title = "kNN (3 neighbor)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")
```

```{r}
# Visualize the KNN classifier with penguins dataset
# 9 neighbors

model <- x |> caret::knn3(species ~ ., data = _, k = 9)
decisionplot(model, x, class_var = "species") + 
  labs(title = "kNN (9 neighbor)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")
```

## Naive Bayes Classifier

```{r}
# Visualize the Naive Bayes classifier with penguins dataset

model <- x |> e1071::naiveBayes(species ~ ., data = _)
decisionplot(model, x, class_var = "species", 
             predict_type = c("class", "raw")) + 
  labs(title = "Naive Bayes",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction") 
```

## Linear Discriminant Analysis

```{r}
# Visualize the Linear Discriminant Analysis with penguins dataset

model <- x |> MASS::lda(species ~ ., data = _)
decisionplot(model, x, class_var = "species") + 
  labs(title = "LDA",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")
```

## Multinomial Logistic Regression

```{r}
# Multinomial Logistic Regression implemented in nnet
model <- x |> nnet::multinom(species ~., data = _)
```

```{r}
# Plotting the model
decisionplot(model, x, class_var = "species") + 
  labs(title = "Multinomial Logistic Regression",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")
```

## Decision Trees

```{r}
# Visualize the CART Decision Tree with penguins dataset

model <- x |> rpart::rpart(species ~ ., data = _)
decisionplot(model, x, class_var = "species") + 
  labs(title = "CART",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")
```

```{r}
# Create an overfitting CART model
model <- x |> rpart::rpart(species ~ ., data = _,
  control = rpart.control(cp = 0.001, minsplit = 1))
decisionplot(model, x, class_var = "species") + 
  labs(title = "CART (overfitting)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")
```

```{r}
# C5.0 Model
model <- x |> C50::C5.0(species ~ ., data = _)
decisionplot(model, x, class_var = "species") + 
  labs(title = "C5.0",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")
```

```{r}
# Using Random Forest model

model <- x |> randomForest::randomForest(species ~ ., data = _)
decisionplot(model, x, class_var = "species") + 
  labs(title = "Random Forest",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")
```

## Support Vector Machine

```{r}
# Visualize the Support Vector Machine model with penguins dataset
# Linear Kernel

model <- x |> e1071::svm(species ~ ., data = _, kernel = "linear")
decisionplot(model, x, class_var = "species") + 
  labs(title = "SVM (linear kernel)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")
```

```{r}
# Visualize the Support Vector Machine model with penguins dataset
# Radial Kernel

model <- x |> e1071::svm(species ~ ., data = _, kernel = "radial")
decisionplot(model, x, class_var = "species") + 
  labs(title = "SVM (radial kernel)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")
```

```{r}

# Visualize the Support Vector Machine model with penguins dataset
# Polynomial Kernel

model <- x |> e1071::svm(species ~ ., data = _, kernel = "polynomial")
decisionplot(model, x, class_var = "species") + 
  labs(title = "SVM (polynomial kernel)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")
```

```{r}
# Visualize the Support Vector Machine model with penguins dataset
# Sigmoid Kernel
model <- x |> e1071::svm(species ~ ., data = _, kernel = "sigmoid")
decisionplot(model, x, class_var = "species") + 
  labs(title = "SVM (sigmoid kernel)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")
```

## Single Layer Feed-forward Neural Networks

```{r}
# Visualize the Neural network model with penguins dataset
# 1 neuron
model <-x |> nnet::nnet(species ~ ., data = _, size = 1, trace = FALSE)
decisionplot(model, x, class_var  = "species", 
  predict_type = c("class", "raw")) + 
  labs(title = "NN (1 neuron)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")
```

```{r}
# Visualize the Neural network model with penguins dataset
# 2 neurons
model <-x |> nnet::nnet(species ~ ., data = _, size = 2, trace = FALSE)
decisionplot(model, x, class_var  = "species", 
  predict_type = c("class", "raw")) + 
  labs(title = "NN (2 neurons)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")
```

```{r}
# Visualize the Neural network model with penguins dataset
# 4 neurons
model <-x |> nnet::nnet(species ~ ., data = _, size = 4, trace = FALSE)
decisionplot(model, x, class_var  = "species", 
  predict_type = c("class", "raw")) + 
  labs(title = "NN (4 neurons)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")
```

```{r}
# Visualize the Neural network model with penguins dataset
# 10 neurons
model <-x |> nnet::nnet(species ~ ., data = _, size = 10, trace = FALSE)
decisionplot(model, x, class_var  = "species", 
  predict_type = c("class", "raw")) + 
  labs(title = "NN (10 neurons)",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       shape = "Species",
       fill = "Prediction")
```

# Circle Dataset

```{r}

set.seed(1000)

x <- mlbench::mlbench.circle(500)
x <- cbind(as.data.frame(x$x), factor(x$classes))
colnames(x) <- c("x", "y", "class")
x <- as_tibble(x)
x
```

```{r}
# Plotting the class variable
ggplot(x, aes(x = x, y = y, color = class)) + 
  geom_point() +
  theme_minimal(base_size = 14)
```

## K-Nearest Neighbors Classifier

```{r}
# Visualize the KNN classifier with circle dataset
# 1 neighbor
model <- x |> caret::knn3(class ~ ., data = _, k = 1)
decisionplot(model, x, class_var = "class") + 
  labs(title = "kNN (1 neighbor)",
       shape = "Class",
       fill = "Prediction")
```

```{r}
# Visualize the KNN classifier with circle dataset
# 10 neighbors
model <- x |> caret::knn3(class ~ ., data = _, k = 10)
decisionplot(model, x, class_var = "class") + 
  labs(title = "kNN (10 neighbors)",
       shape = "Class",
       fill = "Prediction")
```

## Naive Bayes Classifier

```{r}
# Visualize the Naive Bayes classifier with circle dataset

model <- x |> e1071::naiveBayes(class ~ ., data = _)
decisionplot(model, x, class_var = "class", 
  predict_type = c("class", "raw")) + 
  labs(title = "Naive Bayes",
       shape = "Class",
       fill = "Prediction")
```

## Linear Discriminant Analysis

```{r}
# Visualize the Linear Discriminant Analysis with circle dataset
# Not a good model since the true decision boundary is not linear
model <- x |> MASS::lda(class ~ ., data = _)
decisionplot(model, x, class_var = "class") + 
  labs(title = "LDA",
       shape = "Class",
       fill = "Prediction")
```

## Logistic Regression

```{r}
# Create a logistic regression model implemented in nnet
model <- x |> nnet::multinom(class ~., data = _)
```

```{r}
# Plotting the model
decisionplot(model, x, class_var = "class") + 
  labs(title = "Multinomial Logistic Regression",
       shape = "Class",
       fill = "Prediction")
```

## Decision Trees

```{r}
# Visualize the CART model with circle dataset

model <- x |> rpart::rpart(class ~ ., data = _)
decisionplot(model, x, class_var = "class") + 
  labs(title = "CART",
       shape = "Class",
       fill = "Prediction")
```

```{r}
# Visualize the CART model with circle dataset
# Overfitting CART model

model <- x |> rpart::rpart(class ~ ., data = _,
  control = rpart.control(cp = 0.001, minsplit = 1))
decisionplot(model, x, class_var = "class") + 
  labs(title = "CART (overfitting)",
       shape = "Class",
       fill = "Prediction")
```

```{r}
# Visualize the C5.0 model with circle dataset

model <- x |> C50::C5.0(class ~ ., data = _)
decisionplot(model, x, class_var = "class") + 
  labs(title = "C5.0",
       shape = "Class",
       fill = "Prediction")
```

```{r}
# Random forest
library(randomForest)
model <- x |> randomForest(class ~ ., data = _)
decisionplot(model, x, class_var = "class") + 
  labs(title = "Random Forest",
       shape = "Class",
       fill = "Prediction")
```

```{r}
# Visualize the Support Vector Machine model with circle dataset
# Linear SVM
model <- x |> e1071::svm(class ~ ., data = _, kernel = "linear")
decisionplot(model, x, class_var = "class") + 
  labs(title = "SVM (linear kernel)",
       shape = "Class",
       fill = "Prediction")
```

```{r}
# Visualize the Support Vector Machine model with circle dataset
# Radial SVM

model <- x |> e1071::svm(class ~ ., data = _, kernel = "radial")
decisionplot(model, x, class_var = "class") + 
  labs(title = "SVM (radial kernel)",
       shape = "Class",
       fill = "Prediction")
```

```{r}
# Visualize the Support Vector Machine model with circle dataset
# Polynomial SVM
model <- x |> e1071::svm(class ~ ., data = _, kernel = "polynomial")
decisionplot(model, x, class_var = "class") + 
  labs(title = "SVM (polynomial kernel)",
       shape = "Class",
       fill = "Prediction")
```

```{r}
# Visualize the Support Vector Machine model with circle dataset
# Sigmoid SVM
model <- x |> e1071::svm(class ~ ., data = _, kernel = "sigmoid")
decisionplot(model, x, class_var = "class") + 
  labs(title = "SVM (sigmoid kernel)",
       shape = "Class",
       fill = "Prediction")
```

```{r}
# Single Layer Feed-forward Neural Networks
# Visualize the Neural network with circle dataset
# 1 neuron
model <-x |> nnet::nnet(class ~ ., data = _, size = 1, trace = FALSE)
decisionplot(model, x, class_var = "class", 
  predict_type = c("class")) + 
  labs(title = "NN (1 neuron)",
       shape = "Class",
       fill = "Prediction")
```

```{r}
# Single Layer Feed-forward Neural Networks
# Visualize the Neural network with circle dataset
# 2 neurons

model <-x |> nnet::nnet(class ~ ., data = _, size = 2, trace = FALSE)
decisionplot(model, x, class_var = "class", 
  predict_type = c("class")) + 
  labs(title = "NN (2 neurons)",
       shape = "Class",
       fill = "Prediction")
```

```{r}
# Single Layer Feed-forward Neural Networks
# Visualize the Neural network with circle dataset
# 4 neurons
model <-x |> nnet::nnet(class ~ ., data = _, size = 4, trace = FALSE)
decisionplot(model, x, class_var = "class", 
  predict_type = c("class")) + 
  labs(title = "NN (4 neurons)",
       shape = "Class",
       fill = "Prediction")
```

```{r}
# Single Layer Feed-forward Neural Networks
# Visualize the Neural network with circle dataset
# 10 neuron

model <-x |> nnet::nnet(class ~ ., data = _, size = 10, trace = FALSE)
decisionplot(model, x, class_var = "class", 
  predict_type = c("class")) + 
  labs(title = "NN (10 neurons)",
       shape = "Class",
       fill = "Prediction")
```



